---
title: "Homework2"
author: "Haoyuan Dou"
date: "2026-10-06"
output: github_document
---

```{r}
library(tidyverse)
library(janitor)
```

```{r}
# Problem1
subway_df = read.csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv")

subway_df =
  subway_df |>
  clean_names() |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entrance_type,
    entry,
    vending,
    ada
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
```

```{r}
names(subway_df)
```

```{r}
class(subway_df$entry)
```
# This data set contains information on NYC subway station entrances and exits, including subway line, station name and location, routes, entrance type, entry status, vending availability and ADA compliance. I cleaned the data by standardizing variable names, selecting the variables relevant to this problem, and converting entry from a character variable to a logical variable. The resulting data set has `r nrow(subway_df)` rows and `r ncol(subway_df)` columns. These data are not fully tidy because route information is stored across multiple variable rather than in a single route variable.

```{r}
# How many distinct stations are there?
subway_df |>
  distinct(line, station_name) |>
  nrow()

# There are 465 distinct subway stations.
```

```{r}
# How many stations are ADA compliant?
subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

# There are 84 ADA compliant stations.
```

```{r}
# What proportion of station entrances / exits without vending allow entrance?
subway_df |>
  filter(vending == "NO") |>
  summarize(
    entry_proportion = mean(entry)
  )

# The proportion of station entrances / exits without vending that allow entrance is 0.377.
```
```{r}
subway_tidy_df = 
  subway_df |>
  mutate(
    across(route1:route11, as.character)
  ) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    values_to = "route_name"
  )

subway_tidy_df = 
  subway_tidy_df |>
  filter(!is.na(route_name))

subway_tidy_df |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  nrow()

subway_tidy_df |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()

# There are 60 distinct stations that serve the A train, and 17 are ADA compliant.
```

